---
title: "data_wrangling_I"
author: "William Anderson"
date: "2022-09-20"
output: github_document
---


This is the repo associated with code for lectures in data wrangling in P8105

```{r setup, echo = FALSE, message = FALSE}

library(tidyverse)

library(readxl)

library(haven)

```


## Data Import CSVs

Let's import data using he 'readr' package

We should always use relative paths shown by starting the file path as data_import_examples instead of typing C:/Users/Will/Documents etc

```{r}

litters_data = read_csv(file = "data_import_examples/FAS_litters.csv")

pups_data = read_csv(file = "data_import_examples/FAS_pups.csv")

litters_data = janitor::clean_names(litters_data)

names(litters_data)

pups_data = janitor::clean_names(pups_data)

```

look up the data

```{r}
litters_data

head(litters_data)
tail(litters_data)

```


```{r, eval = FALSE}

view(litters_data)

```

```{r}

skimr::skim(litters_data)

```

## Explicitly saying what each column variable type should be

```{r, eval = FALSE}

litters_data = read_csv(file = "data_import_examples/FAS_litters.csv",
  col_types = cols(
    Group = col_character(),
    `Litter Number` = col_character(),
    `GD0 weight` = col_double(),
    `GD18 weight` = col_double(),
    `GD of Birth` = col_integer(),
    `Pups born alive` = col_integer(),
    `Pups dead @ birth` = col_integer(),
    `Pups survive` = col_integer()
  )
)
tail(litters_data)


```


## Other file formats

We need to read in an excel spreadsheet, need to load readxl library

```{r}
mlb_df = read_excel("data_import_examples/mlb11.xlsx")

head(mlb_df, 5) 
```


Viewing mlb data frame

```{r, eval = FALSE}

view(mlb_df)

```

```{r}

lotr_words = read_excel("data_import_examples/LotR_Words.xlsx", range = "B3:D6")

```


## Still more file formats

Read in a SAS dataset, need to load haven library first

```{r}

pulse_data = read_sas("data_import_examples/public_pulse_data.sas7bdat")

head(pulse_data, 5)

```

## Data export examples

```{r}

write_csv(lotr_words, file = "data_import_examples/lotr_words_output.csv")


```

## Why not base r??? aka don't do this for importing

```{r}

dont_do_this_df = read.csv("data_import_examples/FAS_litters.csv")

```


